\documentclass[11pt]{report}

\usepackage{packages}

\begin{document}

\input{title.tex}

\frontmatter

\begin{abstract}
Neutron stars contain matter at extremely high densities, where both general relativity and
quantum many-body physics are required to describe their structure. In this thesis, we study
how different models of dense matter affect the relation between the mass and radius of a
neutron star. The starting point is the Tolman--Oppenheimer--Volkoff equation, which we
solve numerically for a sequence of equations of state.

We first consider simple models based on degenerate fermions. These illustrate how the
pressure of a Fermi gas supports a star against gravitational collapse, and why such idealised
descriptions are not sufficient to describe the full range of observed neutron-star properties.
Building on this, the main part of the thesis develops an interacting model for nucleons and
leptons in equilibrium. The interactions are described through self-consistent meson fields
and give rise to a more realistic equation of state, which is combined with a standard crust
model before being used in the numerical stellar calculations.

The resulting mass–radius relation produces reasonable radii for typical neutron-star masses,
but it does not support the heaviest observed stars and shows clear differences from current
astronomical constraints. This demonstrates that while the interacting model represents a
meaningful improvement over simple free-gas descriptions, it remains incomplete at the
highest densities.

Overall, the thesis presents a coherent path from ideal models to an interacting description of
dense matter and shows the need for more advanced equations of state in order to reproduce
the full range of observed neutron-star properties.
\end{abstract}

\tableofcontents

\mainmatter

\input{chapters/introduction.tex}
\input{chapters/tov.tex}
\input{chapters/ideal_fermi_gas.tex}
\input{chapters/ideal_neutron_stars.tex}
\input{chapters/npemu_model.tex}
\input{chapters/conclusion.tex}

\appendix
\input{appendix/gr_basics.tex}

\printbibliography

\end{document}
